---
title: "Final Project"
author: "Shelley Porter"
date: "`r Sys.Date()`"
output: html_document
#output: flexdashboard::flex_dashboard
---
#didn't like output with the flex_board screen formating so I commented out and made it html. I left it in the code so you saw that I did attempt it.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load library
```{r, warning=FALSE}
library(naniar)
library(flextable)
library(tidyverse)
library(stringr)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggbeeswarm)
library(skimr)
library(plotly)
library(DT)
library(scales)
library(boot)

#set theme to bw
theme_set(theme_bw())
```

Data import
```{r}
# Import data 
census_data <- read.csv("C:/Users/spsky/OneDrive/UoM/R/Final Project/census_data_county_2009-2021.cvs")

# first few rows of the data
head(census_data)
```

Exploratory Data Analysis (EDA)
```{r}
# Data skim
skim(census_data)

# Display data structure
str(census_data)

# Summary of important variables
summary(census_data)

# Missing values
gg_miss_var(census_data)
```

String Manipulation
```{r}
# Manipulate strings in the dataset and append new columns with the changes
census_data <- census_data |>
  mutate(
    new_county_state = str_to_title(county_state),
    new_county_state = str_replace_all(new_county_state, " County", ""),
    year = str_trim(as.character(year))
  )

# Display the first few rows of the updated dataset
head(census_data)
```


Data Cleaning
```{r}
census_data_sample <- census_data|>
  filter(!is.na(median_income))|>
  select(new_county_state, year, population, median_income, prop_female, prop_male, prop_poverty, median_monthly_rent_cost, median_monthly_home_cost)

#cleaned and sampled data
head(census_data_sample)
```

Summary statistics
Create a Sample of the Data
```{r}
# Sample of around 500 data observations
census_data_sample <- census_data |> sample_frac(0.012)

# Check the sample size
nrow(census_data_sample)
```

Group-level Summary Statistics
```{r}
# Summary statistics by new_county_state
state_summary <- census_data |>
  group_by(new_county_state) |>
  summarize(
    avg_income = round(mean(median_income, na.rm = TRUE), 0),
    avg_population = round(mean(population, na.rm = TRUE), 0)
  ) |>
  rename(
    `County, State` = new_county_state,
    `Income` = avg_income,
    `Population` = avg_population
  ) |>
  arrange(`Income`)

# Convert income and population formats
state_summary <- state_summary |>
  mutate(
    Income = dollar(Income),
    Population = comma(Population)
  )

# Display the first 50 rows of the summary table
knitr::kable(head(state_summary, 50), caption = "Average County-level Income Statistics (Lowest 50)")
```


Frequency Table of Two Categorical Variables
```{r}
# Frequency table of county_state and year with average median_monthly_rent_cost
state_year_rent_table <- census_data_sample |>
  group_by(new_county_state, year) |>
  summarize(
    avg_median_rent = round(mean(median_monthly_rent_cost, na.rm = TRUE), 0)
  ) |>
  ungroup() |>
  select(`County, State` = new_county_state, Year = year, `Rent` = avg_median_rent) |>
  arrange(desc(Rent)) |>
  mutate(Rent = dollar(Rent))

# Display the first 50 rows of the frequency table
knitr::kable(head(state_year_rent_table, 50), caption = "Frequency Table of County, Year and Average Median Rent \n(Top 50)")
```


Data Visualizations
Visualization 1: Lowest Average Population 
```{r}
# Filter for the 10 counties with the lowest average populations over the years
lowest_10_population_counties <- census_data_sample |>
  group_by(new_county_state) |>
  summarize(avg_population = mean(population, na.rm = TRUE)) |>
  top_n(-10, avg_population)

# Filter the original data to include only these counties
lowest_10_population_data <- census_data_sample |>
  filter(new_county_state %in% lowest_10_population_counties$new_county_state)

# Summarize the average population for each county
lowest_10_population_summary <- lowest_10_population_data |>
  group_by(new_county_state) |>
  summarize(avg_population = mean(population, na.rm = TRUE))

# Create a horizontal bar chart
ggplot(lowest_10_population_summary, aes(x = reorder(new_county_state, avg_population), y = avg_population)) +
  geom_bar(stat = "identity", fill = "lightblue", color = "black") +
  coord_flip() +
  labs(title = "10 Counties with the \nLowest Average Population", x = "County", y = "Population",
       caption = "Source: Census Data Sample") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```


Visualization 2: Median Income Distribution
```{r}
# Filter for the 10 counties with the highest median incomes
top_10_income_counties <- census_data |>
  group_by(new_county_state) |>
  summarize(median_income = mean(median_income, na.rm = TRUE)) |>
  top_n(10, median_income)

# Filter the original data to include only these counties
top_10_income_data <- census_data |>
  filter(new_county_state %in% top_10_income_counties$new_county_state)

# Median income distribution by top 10 counties using census data
ggplot(top_10_income_data, aes(x = new_county_state, y = round(median_income, 0))) +
  geom_boxplot(fill = "yellow", color = "black") +
  scale_y_continuous(labels = dollar_format(prefix = "$", suffix = "", big.mark = ","), limits = c(80000, 160000), breaks = seq(0, 160000, by = 20000)) +
  labs(
    title = "Median Income Distribution in the \n(Top 10 Counties)",
    x = "County",
    y = "Income",
    caption = "Source: Census Data") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  ) +
  coord_flip()
```

Visualization 3: Population of Females by Top 10 States
```{r}
# Filter for the 10 states with the highest average female population
top_10_female_population_states <- census_data |>
  group_by(new_county_state) |>
  summarize(avg_prop_female = mean(prop_female, na.rm = TRUE)) |>
  top_n(10, avg_prop_female)

# Filter the original data to include only these states
top_10_female_population_data <- census_data |>
  filter(new_county_state %in% top_10_female_population_states$new_county_state)

# Create a heatmap of the proportion of females by top 10 states using sample data
ggplot(top_10_female_population_data, aes(x = new_county_state, y = year, fill = round(prop_female * 100, 0))) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Proportion of Females (%)") +
  labs(
    title = "Population of Females by Top 10 States",
    x = "State",
    y = "Year",
    caption = "Source: Census Data"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6),
    plot.title = element_text(size = 10),
    plot.caption = element_text(size = 8)
  )
```

*Interpretation of the data:* We expect that in Norton City, Virginia, in the year 2014, the population of females reached 63%, making it the highest among the top 10 states with the highest average female population. This suggests that you would most likely find a bride in Norton City, Virginia, during that year based on the highest female population proportion.


Population of Males by Top 10 States
```{r}
# Filter for the 10 states with the highest average male population
top_10_male_population_states <- census_data |>
  group_by(new_county_state) |>
  summarize(avg_prop_male = mean(prop_male, na.rm = TRUE)) |>
  top_n(10, avg_prop_male)

# Filter the original data to include only these states
top_10_male_population_data <- census_data |>
  filter(new_county_state %in% top_10_male_population_states$new_county_state)

# Create a heatmap of the proportion of males by top 10 states using sample data
ggplot(top_10_male_population_data, aes(x = new_county_state, y = year, fill = round(prop_male * 100, 0))) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Proportion of Males (%)") +
 
  labs(title = "Population of Males by Top 10 States ", x = "State", y = "Year",
caption = "Source: Cenus Data") +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6),
    plot.title = element_text(size = 10),
    plot.caption = element_text(size = 8)
  )
```

*Interpretation of the data:* In 2017, Forest, Pennsylvania had the highest population of males, reaching up to 80%. This makes it the state and year where you would most likely find a husband based on the highest male population.


Visualization 4: Population in Poverty Trend
```{r}
lowest_10_poverty_counties <- census_data_sample |>
  group_by(new_county_state) |>
  summarize(avg_prop_poverty = mean(prop_poverty, na.rm = TRUE)) |>
  top_n(-10, avg_prop_poverty)

# Filter the original data to include only these counties
lowest_10_poverty_data <- census_data_sample |>
  filter(new_county_state %in% lowest_10_poverty_counties$new_county_state)

# Plot the trend of poverty rates over time for the lowest 10 counties
ggplot(lowest_10_poverty_data, aes(x = year, y = prop_poverty * 100, color = new_county_state, group = new_county_state)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Trend of Poverty Rates Over Time for the \n10 Counties with the Lowest Average Poverty Rates",
    x = "Year",
    y = "Population in Poverty",
    color = "County",
    caption = "Source: Census Data Sample"
  )+
  theme(axis.text.y = element_text(size = 7),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```


Visualization 5: Interactive comparison of Median income level with Median Rent and Mortgage Costs
```{r}
# Convert 'year' to numeric
census_data <- census_data |> mutate(year = as.numeric(year))

# Summary of median income, rent cost, and home cost
summary_data <- census_data |>
  group_by(year) |>
  summarise(
     mean_income = round(mean(median_income, na.rm = TRUE), 2),
    mean_rent_cost = round(mean(median_monthly_rent_cost, na.rm = TRUE), 2),
    mean_home_cost = round(mean(median_monthly_home_cost, na.rm = TRUE), 2)
  )

# Melt the data for faceting
summary_data_long <- summary_data |>
  pivot_longer(cols = c(mean_income, mean_rent_cost, mean_home_cost),
               names_to = "variable",
               values_to = "value")

# Recode the variable names
summary_data_long <- summary_data_long |>
  mutate(variable = recode(variable,
                           mean_income = "Income",
                           mean_rent_cost = "Rent",
                           mean_home_cost = "Mortgage"))

# Plot the trends
comparison_plot <- ggplot(summary_data_long, aes(x = year, y = value, color = variable)) +
  geom_line(size = 1.2) +
  scale_y_continuous(labels = dollar_format(prefix = "$", suffix = "", big.mark = ",", accuracy = 0.01)) +
   facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  labs(
    title = 'Comparison of Income, Rent and Mortgage Cost \n(2009-2021)',
    x = 'Year',
    y = 'Amount',
    caption = "Data Source: United States Census Data",
    color = "Variable"
  )
# Convert ggplot to plotly
ggplotly(comparison_plot)
```

Beeswarm Plot
```{r}
# Filter counties with median income between 120,000 and 160,000
income_filtered <- census_data|>
  filter(median_income >= 120000 & median_income <= 160000)

# beeswarm plot
ggplot(income_filtered, aes(x = reorder(new_county_state, -median_income), y = median_income)) +
  geom_quasirandom(alpha = 0.6, color = "blue") +
  labs(title = "Beeswarm Plot of Median Income by County\n(120,000 and 160,000)",
       x = "County",
       y = "Income",
caption = "Source: Cenus Data") +
  scale_y_continuous(labels = dollar_format(prefix = "$", suffix = "", big.mark = ","), breaks = seq(120000, 160000, by = 15000)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  ) +
  guides(color=guide_legend(title="County"))
```



Permutation Testing
```{r}
# Permutation test for difference in median income between two counties
county1 <- "Antrim County, Michigan"
county2 <- "Wayne County, Michigan"

# Define the function to compute income difference
income_diff <- function(data, county1, county2) {
  mean(data[data$county_state == county1, ]$median_income, na.rm = TRUE) - 
  mean(data[data$county_state == county2, ]$median_income, na.rm = TRUE)
}

# Calculate the observed difference
observed_diff <- income_diff(census_data, county1, county2)

# Generate permuted differences
permuted_diffs <- replicate(1000, {
  permuted_data <- census_data
  permuted_data$county_state <- sample(permuted_data$county_state)
  income_diff(permuted_data, county1, county2)
})

# Convert the permuted differences to a data frame
permuted_diffs_df <- data.frame(permuted_diffs = permuted_diffs)

# Plot the permuted differences
ggplot(permuted_diffs_df, aes(x = permuted_diffs)) +
  geom_histogram(binwidth = 500, fill = "blue", alpha = 0.7, color = "black") +
  geom_vline(aes(xintercept = observed_diff), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Permutation Test for Median Income Difference", x = "Income Difference", y = "Frequency",
caption = "Source: Cenus Data") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  ) 
```

Bootstrap
```{r}
# Set seed for reproducibility
set.seed(1971)

# Number of bootstrap samples
B <- 1000

# Instantiating matrix for bootstrap samples
bootstrap_samples <- matrix(NA, nrow = nrow(census_data_sample), ncol = B)

# Sampling with replacement B times
for (b in 1:B) {
  bootstrap_samples[, b] <- census_data_sample |>
    slice_sample(prop = 1, replace = TRUE) |>
    dplyr::pull(median_income)
}

# Calculate bootstrap medians
bootstrap_medians <- apply(bootstrap_samples, 2, median, na.rm = TRUE)

# Calculate bootstrap confidence interval
boot_ci <- quantile(bootstrap_medians, probs = c(0.025, 0.975))

# Create a histogram of the bootstrap medians
ggplot(data.frame(medians = bootstrap_medians), aes(x = round(medians, 0))) +
  geom_histogram(binwidth = 500) +
  geom_vline(aes(xintercept = round(boot_ci[1], 0)), color = "red") +
  geom_vline(aes(xintercept = round(boot_ci[2], 0)), color = "red") +
  labs(
    title = "Bootstrap Confidence Interval for Median Income",
    x = "Income",
    y = "Frequency",
    caption = "Source: Census Data Sample"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```

Data Dictionary
```{r}
# Display data dictionary
data_dictionary <- tibble(
  variable = c("geoid", "county_state", "year", "population", "median_income", 
               "median_monthly_rent_cost", "median_monthly_home_cost", "prop_female", 
               "prop_male", "prop_poverty"),
  description = c("Geographic region ID", "Geographic region", "Year", "Population", 
                  "Median income in dollars", "Median monthly housing costs for homeowners in dollars", 
                  "Median monthly rent costs for renters in dollars", "Proportion of people who are female", 
                  "Proportion of people who are male", "Proportion of people living in poverty")
)

knitr::kable(data_dictionary, caption = "Data Dictionary for Census Data")
```



Extra Credit
Population Trends
```{r}
# Filter data for a sample county
sample_county <- census_data |> dplyr::filter(new_county_state == 'Autauga, Alabama')

# Check the structure of sample_county to ensure it's correct
str(sample_county)

# Plot population trend
plot <- ggplot(sample_county, aes(x = year, y = population)) +
  geom_line(size = 1.2, color = "blue") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = 'Population Trend in Autauga County, Alabama',
    x = 'Year',
    y = 'Population',
    caption = "Data Source: Census Data"
  )

# Convert ggplot to plotly
ggplotly(plot)
```

Median Income Summary
```{r}
# Convert 'year' back to numeric
census_data <- census_data |> mutate(year = as.numeric(year))

# Summary of median income
income_summary <- census_data |>
  group_by(year) |>
  summarise(mean_income = mean(median_income, na.rm = TRUE))

# Check the summary
print(income_summary)

# Plot median income trend
income_plot <- ggplot(income_summary, aes(x = year, y = mean_income)) +
  geom_line(size = 1.2, color="blue") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = 'Median Income Trend (2009-2021)',
    x = 'Year',
    y = 'Median Income (in USD)',
    caption = "Data Source: United States Census Data"
  )

# Convert ggplot to plotly
ggplotly(income_plot)
```


Median Monthly Rent Cost Summary
```{r}
# Summary of median monthly rent cost
rent_summary <- census_data |>
  group_by(year) |>
  summarise(mean_rent = mean(median_monthly_rent_cost, na.rm = TRUE))

# Plot median monthly rent cost trend
rent_plot <- ggplot(rent_summary, aes(x = year, y = mean_rent)) +
  geom_line(size = 1.2, color="blue") +
  labs(
    title = 'Median Monthly Rent Cost Trend (2009-2021)',
    x = 'Year',
    y = 'Monthly Rent',
    caption = "Data Source: Census Data")

# Convert ggplot to plotly
    ggplotly(rent_plot)
```

Key Summary Table
```{r}
# Summary table of key statistics
key_summary <- census_data |>
  group_by(year) |>
  summarise(
    mean_population = mean(population, na.rm = TRUE),
    mean_income = mean(median_income, na.rm = TRUE),
    mean_rent = mean(median_monthly_rent_cost, na.rm = TRUE),
    mean_home_cost = mean(median_monthly_home_cost, na.rm = TRUE)
  ) |>
  mutate(
    mean_population = round(mean_population),
    mean_income = round(mean_income),
    mean_rent = round(mean_rent),
    mean_home_cost = round(mean_home_cost)
  )

# Display the summary table
datatable(key_summary, options = list(pageLength = 10)) |>
  formatCurrency(columns = c('mean_income', 'mean_rent', 'mean_home_cost'), currency = '', interval = 3, mark = ',') |>
  formatRound(columns = 'mean_population', interval = 3, mark = ',', digits = 0)
```

Key Findings:

- Income Distribution: Higher-income counties tend to be located in more economically developed areas, while lower-income counties are often found in less developed regions.

- Rent and Mortgage Trends: Rent and mortgage costs have generally increased over time, reflecting economic inflation and housing market trends.

- Population Dynamics: Certain counties show significant gender imbalances, which could be due to various socio-economic factors.

- Poverty Rates: Counties with the lowest poverty rates consistently show a decline, indicating economic improvements.

The project offers a thorough analysis and visual representations that provide valuable insights into socio-economic trends across various counties in the United States during the period under examination.

